\chapter{Introduction}

\section{Motivation}
Parser combinators \cite{hutton_higher-order_1992} are an elegant approach for writing parsers in a manner which remains close to the original grammar specification.
\texttt{parsley} \cite{willis_garnishing_2018} is a parser combinator library implemented as an embedded domain-specific language (\textsc{dsl}) \cite{hudak_building_1996} in Scala, with an API inspired by the \texttt{parsec} \cite{leijen_parsec_2001} family of libraries in Haskell.
However, as with many libraries, there exists a learning curve to utilising \texttt{parsley} and parser combinator libraries in an idiomatic manner.

While well-documented, the wealth of information to get started with \texttt{parsley} can be overwhelming for users, particularly those new to parser combinators.
Furthermore, there exists a number of design patterns \cite{willis_design_2022} for writing maintainable parsers, which even experienced users may be unaware of.
A potential solution to this problem is tooling to provide automated code hints, which a user can use during the development cycle to evaluate if their code adheres to best practices.

% TODO: mention pitfalls of e.g. left recursion here, or in the background? Stuff which can't be picked up by a regular linter

A number of modern integrated development environments (\textsc{ide}s) provide code hints to warn programmers about problems in their source code, highlighting offending snippets and suggesting actions to improve suboptimal or incorrect code \cite{kurbatova_intellij_2021}.
Many of these code analysis tools are designed to detect general issues for the host language, rather than specifically for libraries.
However, tools may also utilise domain-specific code analyses in order to detect issues specific to a particular system or problem domain \cite{renggli_domain-specific_2010,gregor_stllint_2006,xunit_xunitanalyzers_2024}.

This project aims to explore the potential of harnessing code analysis techniques to develop a new tool, \texttt{parsley-garnish}, that offers code hints aimed at assisting programmers in writing idiomatic and correct \texttt{parsley} code.
Additionally, for certain hints which can be automatically fixed, \texttt{parsley-garnish} will provide automated actions to resolve the issue. % TODO: via code transformations - put this in the background?
The goal of \texttt{parsley-garnish} is to be used as a companion library to \texttt{parsley}, in order to improve its ease of adoption and to help users enforce best practices.

\section{Planned Objectives}
